R-LIME: LIME法における矩形制約と最適化

### 2024-03-15 paper/paper.tex
Work on the paper

% settings for cleveref
\crefname{figure}{Figure}{Figures}
\crefname{table}{Table}{Tables}
\crefname{algorithm}{Algorithm}{Algorithms}

@@

\ifnum\mode=0
  \pagestyle{plain}
  \pagenumbering{roman}
  \tableofcontents
  \newpage
  \listoffigures\listoftables\clearpage
  \AddToHook{cmd/section/before}{\clearpage}
  \pagestyle{plain}
  \pagenumbering{arabic}
\else

  % The abstract should briefly summarize the contents of the paper in
  % 150--250 words.
  \begin{abstract}
    In recent years,
    complex machine learning models such as deep neural networks
    have been used in various industrial fields due to their high accuracy.
    However,
    its complexity has been a major obstacle to implementation
    in decision-making situations where transparency of the decision process
    is required.
    In order to address this problem,
    various post-hoc explanation methods have been proposed,
    but they have not been able to achieve interpretability of
    both the explanation and its scope.
    We propose a new method, R-LIME,
    which interprets complex classifiers in an interpretable scope.
    R-LIME locally approximizes a complex decision boundary linearly
    in a rectangular region
    and maximizes the region as long as the accuracy of the linear
    classifier is higher than a specified threshold.
    The resulting rectangular region is interpretable for users because it is
    expressed as a conjunction of feature predicates.
    We demonstrate the effectiveness of the proposed method through qualitative
    and quantitative experiments using a real-world dataset.
    Finally, we discuss limitations and future work of the proposed method.
    % 148 words
    \keywords{Interpretable machine learning \and Local surrogate model}
  \end{abstract}
\fi

\section{Introduction}
Machine learning models, such as deep learning and random forests,
have been widely employed in various industrial applications
due to their significant improvement in accuracy in recent years.
However,
the increasing complexity and black-box nature of these models pose challenges,
particularly in critical decision-making scenarios like healthcare and finance,
where the opacity of decision process becomes a major obstacle to
implementation.
Consequently,
there has been extensive research in the field of post-hoc explanations
for machine learning models~\cite{%
  guidotti2018local,radulovic2023bella,ribeiro2016why,ribeiro2018anchors}.
Existing post-hoc explanation methods are categorized
into \emph{model-dependent} and \emph{model-agnostic} methods
based on their dependence on the model's structure.
Furthermore, model-agnostic methods are classified into
\emph{global} and \emph{local} methods based on their locality in input space.

This paper specifically focuses
on \emph{model-agnostic} and \emph{local} methods,
which do not rely on the models' structure
and explain the models' behavior
in the vicinity of a given input (focal point).
For example, LIME (Local Interpretable Model-agnostic Explanations)
\cite{ribeiro2016why} approximates the complex model's decision boundary
locally using a simpler model around the given focal point.
Although LIME enables users to interpret the local behavior of complex models,
it does not explicitly provide the effective scope of the explanation
\cite{ribeiro2018anchors}.
This limitation can lead
users to overgeneralize the knowledge derived from the explanation.
Without a clear indication of the scope of the explanation,
users might mistakenly apply
the derived knowledge to instances outside this scope,
potentially leading to misunderstanding of the black-box model's behavior.

@@

\begin{figure}[tbp]
  \def\scale{0.38}
  \centering
  \begin{subfigure}[t]{0.55\textwidth}
    \centering
    \includegraphics[scale=\scale]{example-instance}
    \caption{Focal point.}\label{fig:example-instance}
    \vspace{0.5cm}
  \end{subfigure}
  \begin{subfigure}[t]{0.45\textwidth}
    \begin{subfigure}[t]{\textwidth}
      \centering
      \includegraphics[scale=\scale]{example-lime}
      \caption{%
        Explanation by LIME\@.
        It provides contributions of each feature to the output,
        but does not explicitly indicate its scope.
      }\label{fig:example-lime}
      \vspace{0.4cm}
    \end{subfigure}
    \begin{subfigure}[t]{\textwidth}
      \centering
      \includegraphics[scale=\scale]{example-anchor}
      \caption{%
        Explanation by Anchor.
        It provides the effective scope of the explanation,
        but does not show the influence of each feature.
      }\label{fig:example-anchor}
    \end{subfigure}
  \end{subfigure}
  \hspace{0.3cm}
  \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \ifnum\mode=0
      \vspace{-1.79cm}
    \else
      \vspace{-2.24cm}
    \fi
    \includegraphics[scale=\scale]{example-rlime}
    \caption{%
      Explanation by R-LIME\@.
      It provides both contributions of each feature and
      its effective scope.
    }\label{fig:example-rlime}
  \end{subfigure}
  \caption[Example of explanations by LIME, Anchor and R-LIME]{%
    Example of explanations by LIME~\cite{ribeiro2016why},
    Anchor~\cite{ribeiro2018anchors} and R-LIME (our proposed method).
    for a sentiment prediction model.
  }\label{fig:example}
\end{figure}
In this paper,
we focus on local model-agnostic methods.
An example of explanations by LIME and Anchor
for a sentiment prediction model is illustrated in \cref{fig:example}.
LIME~\cite{ribeiro2016why} approximates the complex decision boundary
locally using a linear model around the given focal point,
then provides the weights of the linear model as the contribution of each feature
to the output.
The explanation by LIME (\cref{fig:example-lime}) provides suggests that
the word ``not'' mainly contributes to the positive prediction,
but does not explicitly indicate the effective scope.
Without the scope,
users might mistakenly apply the knowledge derived from the explanation
to other instances far from the focal point,
potentially leading to misunderstanding of the black-box model's behavior
\cite{ribeiro2018anchors}.
For example,
user may apply the insights derived from the explanation
to the sentence ``This book is not good.''
and mistakenly conclude that the word ``not''
mainly contributes to the positive prediction for this sentence as well,
which is incorrect obviously.
Anchor~\cite{ribeiro2018anchors} represents a rectangular region
containing the focal point,
that maximizes the probability of the black-box classifier outputting
the same label as the focal point.
While Anchor provides an effective scope of the explanation,
the information it includes is less than LIME,
limiting its utility~\cite{ribeiro2018anchors}.
The explanation by Anchor (\cref{fig:example-anchor})
suggests that changing words other than ``not'' and ``bad''
has little impact on the classifier's output.
While it clearly cannot be applied to the sentence ``This book is not good''
because of not including the word ``bad'',
the explanation does not provide details about the influence of each word,
resulting in less user insight into the model's behavior compared to LIME.

% BELLA\cite{radulovic2023bella} addresses the non-optimization of the approximation region,
% a drawback of LIME,
% by learning a local linear classifier $g$
% using a subset $Z'$ of the dataset $Z$
% instead of perturbed samples.
% The process involves:
% \begin{enumerate}
% 	\item Computing the distance $d_x(z)$ for all instances $z$ in $Z$
% 	      with respect to the focal point $x$.
% 	\item Selecting the $k$ instances with the smallest $d_x(z)$,
% 	      forming a subset $Z_s(k) \subseteq Z$.
% 	\item Learning the local linear classifier $g$ using $Z_s(k)$
% 	      and calculating the similarity measure $\mathcal{R}(k)$ between $f$ and $g$
% 	      over $k=1,2,\dots,|Z|$.
% 	\item Obtaining the optimal $k^*$ by maximizing $\mathcal{R}(k)$.
% \end{enumerate}
% 
% BELLA aims to enhance the reliability of explanations
% by optimizing the approximation region for high-accuracy linear approximations.
% However, it assumes direct access to the dataset $Z$ used for training the classifier,
% making it inapplicable in scenarios where direct access to the dataset is restricted.
% Additionally,
% as BELLA presents the approximation region as a subset of the dataset,
% its interpretability is lower compared to Anchor.

\section{Proposed Method}
\begin{figure}[tbp]
  \centering
  \begin{subfigure}[t]{0.3\textwidth}
    \centering
    \includegraphics[width=\textwidth]{visual-lime}
    \caption{%
      LIME: Locally approximates the decision boundary around the focal point.
    }\label{fig:lime}
  \end{subfigure}%
  \hspace{0.03\textwidth}
  \begin{subfigure}[t]{0.3\textwidth}
    \centering
    \includegraphics[width=\textwidth]{visual-anchor}
    \caption{%
      Anchor: Maximizes coverage of a rectangular region
      containing the focal point under accuracy constraints.
    }\label{fig:anchor}
  \end{subfigure}
  \hspace{0.03\textwidth}
  \begin{subfigure}[t]{0.3\textwidth}
    \centering
    \includegraphics[width=\textwidth]{visual-rlime3}
    \caption{%
      R-LIME:
      Maximizes coverage of a rectangular region
      under lower constraints on the accuracy of the linear classifier.
    }\label{fig:rlime}
  \end{subfigure}
  \caption[Visual comparison of LIME, Anchor and R-LIME]{%
    Visual comparison of LIME, Anchor and R-LIME (our proposed method).
    The solid line represents the rectangular region containing the focal
    point, and the dashed line represents the learned approximation model.
  }
\end{figure}
\subsection{Previous Work}
We specifically focuses on \emph{local} and \emph{model-agnostic} methods.
In this section,
we briefly review existing research on local model-agnostic explanations,
particularly focusing on studies closely related to our proposed method.
\subsubsection[LIME]{%
  LIME (Local Interpretable Model-agnostic Explanations)
  \cite{ribeiro2016why}
}
LIME locally
approximates a trained black-box classifier $f: \mathbb{R}^m \to \{0,1\}$
around a focal point $x \in \mathbb{R}^m$
by a linear classifier $g: \mathbb{R}^m \to \{0,1\}$
(\cref{fig:lime}).
The process involves:
\begin{enumerate}
  \item Obtaining a set of perturbed samples $\mathcal{Z}_p$ around $x$
        and the set of pseudo-labels $f(\mathcal{Z}_p) = \{f(z) \mid z \in \mathcal{Z}_p\}$.
  \item Learning a linear classifier $g$
        using $\mathcal{Z}_p$ and $f(\mathcal{Z}_p)$
        by minimizing the following loss function:
        \begin{equation}
          \label{eq:lime_loss}
          \mathcal{L}(f,g,\pi_x)=\sum_{z\in\mathcal{Z}_p}
          \pi_x(z){\left(f(z)-g(z)\right)}^2,
        \end{equation}
        where $\pi_x(z)$ is a weight function designed to be larger for samples
        closer to $x$, typically implemented using an exponential kernel.
\end{enumerate}
LIME provides valuable insights into the local behavior of the model
by showing the contribution of each feature to the output $f(x)$.
However, due to not explicitly indicating the perturbation region,
users cannot assess the effective scope of the explanation
\cite{ribeiro2018anchors}.

@@

  \foreach \alg in {2,3,4}{%
      \begin{algorithm}[p]
        \small
        \input{tex/alg\alg}
      \end{algorithm}
    }
}

The algorithm of our method
is mainly based on that used in Anchor\cite{ribeiro2018anchors}.
For non-convex optimization problems like \cref{eq:main_problem},
greedy search are often used.
However,
greedy methods tend to converge to local optima, and to address this,
R-LIME utilizes beam search,
which selects multiple candidates at each iteration.
The pseudo code is shown in \cref{alg:greedy_search}.

@@

\ifnum\mode=1
  \subsection{Computational Complexity}
  Post-hoc explanation methods including LIME, Anchor, and R-LIME
  need to query the black-box model multiple times
  to get labels for perturbed samples, which is computationally expensive.
  The number of queries required for LIME is $|\mathcal{Z}_p|$,
  where $|\mathcal{Z}_p|$ is the number of perturbed samples.
  On the other hand,
  the number of queries required for Anchor and R-LIME is
  $\mathcal{O}(B\cdot m^2+m^2\cdot\mathcal{O}_{\mathrm{MAB}[B\cdot m,B]})$,
  where $B$ is the beam width, $m$ is the number of features, and
  $\mathcal{O}_{\mathrm{MAB}[B\cdot m,B]}$ is the computational complexity of
  pure exploration problem of finding the $B$ best arms among $B\cdot m$ arms.
  $\mathcal{O}_{\mathrm{MAB}[B\cdot m,B]}$ depends on hyperparameters
  such as threshold $\tau$, confidence level $1-\delta$ and tolerance $\epsilon$.
\fi

>>>>

% settings for cleveref
\crefname{table}{Table}{Tables}
\crefname{algorithm}{Algorithm}{Algorithms}
\ifnum\mode=0
  \crefname{figure}{Figure}{Figures}
\else
  \crefname{figure}{Fig.}{Figs.}
\fi

@@

\ifnum\mode=0
  \pagestyle{plain}
  \pagenumbering{roman}
  \tableofcontents
  \newpage
  \listoffigures\listoftables\clearpage
  \AddToHook{cmd/section/before}{\clearpage}
  \pagestyle{plain}
  \pagenumbering{arabic}
\else
  % The abstract should briefly summarize the contents of the paper in
  % 150--250 words.
  \begin{abstract}
    In recent years,
    complex machine learning models such as deep neural networks
    have been used in various industrial fields due to their high accuracy.
    However,
    its complexity has been a major obstacle to implementation
    in decision-making situations where transparency of the decision process
    is required.
    In order to address this problem,
    various post-hoc explanation methods have been proposed,
    but they have not been able to achieve interpretability of
    both the explanation and its scope.
    We propose a new method, R-LIME,
    which interprets complex classifiers in an interpretable scope.
    R-LIME locally approximizes a complex decision boundary linearly
    in a rectangular region
    and maximizes the region as long as the accuracy of the linear
    classifier is higher than a specified threshold.
    The resulting rectangular region is interpretable for users because it is
    expressed as a conjunction of feature predicates.
    We demonstrate the effectiveness of the proposed method through qualitative
    and quantitative experiments using a real-world dataset.
    Finally, we discuss limitations and future work of the proposed method.
    % 148 words
    \keywords{Interpretable machine learning \and Local surrogate model}
  \end{abstract}
\fi

\section{Introduction}
Machine learning models, such as deep learning and random forests,
have been widely employed in various industrial applications
due to their significant improvement in accuracy in recent years.
However,
the increasing complexity and black-box nature of these models pose challenges,
particularly in critical decision-making scenarios like healthcare and finance,
where the opacity of decision process becomes a major obstacle to
implementation.
Consequently,
there has been extensive research in the field of post-hoc explanations
for machine learning models~\cite{%
  guidotti2018local,radulovic2023bella,ribeiro2016why,ribeiro2018anchors}.
Existing post-hoc explanation methods are categorized
into \emph{model-dependent} and \emph{model-agnostic} methods
based on their dependence on the model's structure.
Furthermore, model-agnostic methods are classified into
\emph{global} and \emph{local} methods based on their locality in input space.
% This paper specifically focuses
% on \emph{model-agnostic} and \emph{local} methods,
% which do not rely on the models' structure
% and explain the models' behavior
% in the vicinity of a given input (focal point).
% For example, LIME (Local Interpretable Model-agnostic Explanations)
% \cite{ribeiro2016why} approximates the complex model's decision boundary
% locally using a simpler model around the given focal point.
% Although LIME enables users to interpret the local behavior of complex models,
% it does not explicitly provide the effective scope of the explanation
% \cite{ribeiro2018anchors}.
% This limitation can lead
% users to overgeneralize the knowledge derived from the explanation.
% Without a clear indication of the scope of the explanation,
% users might mistakenly apply
% the derived knowledge to instances outside this scope,
% potentially leading to misunderstanding of the black-box model's behavior.
  \def\scale{0.38}
  \begin{subfigure}[t]{0.55\textwidth}
    \includegraphics[scale=\scale]{example-instance}
    \caption{Focal point.}\label{fig:example-instance}
    \vspace{0.5cm}
  \begin{subfigure}[t]{0.45\textwidth}
    \begin{subfigure}[t]{\textwidth}
      \includegraphics[scale=\scale]{example-lime}
      \caption{%
        Explanation by LIME\@.
        It provides contributions of each feature to the output,
        but does not explicitly indicate its scope.
      }\label{fig:example-lime}
      \vspace{0.4cm}
    \begin{subfigure}[t]{\textwidth}
      \includegraphics[scale=\scale]{example-anchor}
      \caption{%
        Explanation by Anchor.
        It provides the effective scope of the explanation,
        but does not show the influence of each feature.
      }\label{fig:example-anchor}
  \hspace{0.3cm}
  \begin{subfigure}[t]{0.45\textwidth}
      \vspace{-1.79cm}
      \vspace{-2.24cm}
    \includegraphics[scale=\scale]{example-rlime}
    \caption{%
      Explanation by R-LIME\@.
      It provides both contributions of each feature and
      its effective scope.
    }\label{fig:example-rlime}
  \caption[Example of explanations by LIME, Anchor and R-LIME]{%
    Example of explanations by LIME~\cite{ribeiro2016why},
    Anchor~\cite{ribeiro2018anchors} and R-LIME (our proposed method).
    for a sentiment prediction model.
  }\label{fig:example}
In this paper,
we focus on local model-agnostic methods.
An example of explanations by LIME and Anchor
for a sentiment prediction model is illustrated in \cref{fig:example}.
LIME~\cite{ribeiro2016why} approximates the complex decision boundary
locally using a linear model around the given focal point,
then provides the weights of the linear model as the contribution of each feature
to the output.
The explanation by LIME (\cref{fig:example-lime}) provides suggests that
the word ``not'' mainly contributes to the positive prediction,
but does not explicitly indicate the effective scope.
Without the scope,
users might mistakenly apply the knowledge derived from the explanation
to other instances far from the focal point,
potentially leading to misunderstanding of the black-box model's behavior
\cite{ribeiro2018anchors}.
For example,
user may apply the insights derived from the explanation
to the sentence ``This book is not good.''
and mistakenly conclude that the word ``not''
mainly contributes to the positive prediction for this sentence as well,
which is incorrect obviously.
Anchor~\cite{ribeiro2018anchors} represents a rectangular region
containing the focal point,
that maximizes the probability of the black-box classifier outputting
the same label as the focal point.
While Anchor provides an effective scope of the explanation,
the information it includes is less than LIME,
limiting its utility~\cite{ribeiro2018anchors}.
The explanation by Anchor (\cref{fig:example-anchor})
suggests that changing words other than ``not'' and ``bad''
has little impact on the classifier's output.
While it clearly cannot be applied to the sentence ``This book is not good''
because of not including the word ``bad'',
the explanation does not provide details about the influence of each word,
resulting in less user insight into the model's behavior compared to LIME.

@@

\section{Proposed Method}
\begin{figure}[tbp]
  \centering
  \begin{subfigure}[t]{0.3\textwidth}
    \centering
    \includegraphics[width=\textwidth]{visual-lime}
    \caption{%
      LIME: Locally approximates the decision boundary around the focal point.
    }\label{fig:lime}
  \end{subfigure}%
  \hspace{0.03\textwidth}
  \begin{subfigure}[t]{0.3\textwidth}
    \centering
    \includegraphics[width=\textwidth]{visual-anchor}
    \caption{%
      Anchor: Maximizes coverage of a rectangular region
      containing the focal point under accuracy constraints.
    }\label{fig:anchor}
  \end{subfigure}
  \hspace{0.03\textwidth}
  \begin{subfigure}[t]{0.3\textwidth}
    \centering
    \includegraphics[width=\textwidth]{visual-rlime3}
    \caption{%
      R-LIME:
      Maximizes coverage of a rectangular region
      under lower constraints on the accuracy of the linear classifier.
    }\label{fig:rlime}
  \end{subfigure}
  \caption[Visual comparison of LIME, Anchor and R-LIME]{%
    Visual comparison of LIME, Anchor and R-LIME (our proposed method).
    The solid line represents the rectangular region containing the focal
    point, and the dashed line represents the learned approximation model.
  }
\end{figure}

\subsection{Previous Work}
We specifically focuses on \emph{local} and \emph{model-agnostic} methods.
In this section,
we briefly review existing research on local model-agnostic explanations,
particularly focusing on studies closely related to our proposed method.

\subsubsection[LIME]{%
  LIME (Local Interpretable Model-agnostic Explanations)
  \cite{ribeiro2016why}
}
LIME locally
approximates a trained black-box classifier $f: \mathbb{R}^m \to \{0,1\}$
around a focal point $x \in \mathbb{R}^m$
by a linear classifier $g: \mathbb{R}^m \to \{0,1\}$
(\cref{fig:lime}).
The process involves:
\begin{enumerate}
  \item Obtaining a set of perturbed samples $\mathcal{Z}_p$ around $x$
        and the set of pseudo-labels $f(\mathcal{Z}_p) = \{f(z) \mid z \in \mathcal{Z}_p\}$.
  \item Learning a linear classifier $g$
        using $\mathcal{Z}_p$ and $f(\mathcal{Z}_p)$
        by minimizing the following loss function:
        \begin{equation}
          \label{eq:lime_loss}
          \mathcal{L}(f,g,\pi_x)=\sum_{z\in\mathcal{Z}_p}
          \pi_x(z){\left(f(z)-g(z)\right)}^2,
        \end{equation}
        where $\pi_x(z)$ is a weight function designed to be larger for samples
        closer to $x$, typically implemented using an exponential kernel.
\end{enumerate}
LIME provides valuable insights into the local behavior of the model
by showing the contribution of each feature to the output $f(x)$.
However, due to not explicitly indicating the perturbation region,
users cannot assess the effective scope of the explanation
\cite{ribeiro2018anchors}.

@@

  \foreach \alg in {2,3,4}{%
      \begin{algorithm}[p]
        \small
        \input{tex/alg\alg}
      \end{algorithm}
    }
}
The algorithm of our method
is mainly based on that used in Anchor\cite{ribeiro2018anchors}.
For non-convex optimization problems like \cref{eq:main_problem},
greedy search are often used.
However,
greedy methods tend to converge to local optima, and to address this,
R-LIME utilizes beam search,
which selects multiple candidates at each iteration.
The pseudo code is shown in \cref{alg:greedy_search}.

@@

\ifnum\mode=1
  \subsection{Computational Complexity}
  Post-hoc explanation methods including LIME, Anchor, and R-LIME
  need to query the black-box model multiple times
  to get labels for perturbed samples, which is computationally expensive.
  The number of queries required for LIME is $|\mathcal{Z}_p|$,
  where $|\mathcal{Z}_p|$ is the number of perturbed samples selected by the user.
  On the other hand,
  the number of queries required for Anchor and R-LIME is
  $\mathcal{O}(B\cdot m^2+m^2\cdot\mathcal{O}_{\mathrm{MAB}[B\cdot m,B]})$,
  where $B$ is the beam width, $m$ is the number of features, and
  $\mathcal{O}_{\mathrm{MAB}[B\cdot m,B]}$ is the computational complexity of
  pure exploration problem of finding the $B$ best arms among $B\cdot m$ arms.
  $\mathcal{O}_{\mathrm{MAB}[B\cdot m,B]}$ depends on hyperparameters
  such as threshold $\tau$, confidence level $1-\delta$ and tolerance $\epsilon$.
\fi
